---
title: "Homework 4"
output:
  pdf_document: default
---

# Problem 1


Let's first deal with the likelihood.
$$
\begin{array}{l}
Y=X \alpha +\varepsilon \\
\varepsilon=Y- X \alpha \\
\varepsilon_{i}=Y_{i}-X_{i} \alpha
\end{array}
$$

It is given that $\varepsilon$ follows multivariate normal distribution with a diagonal covariance matrix, which means that each $\varepsilon_{i}$ is independent of others. Therefore:

$$
\begin{aligned}
\ell\left(\sigma^{2}\right) &=f\left(\varepsilon | \sigma^{2}\right)=N\left(\varepsilon | 0, \sigma^{2} I_{n}\right) \\
&=\prod_{i=1}^{n} N\left(\varepsilon_{i} | 0, \sigma^{2}\right)=\prod_{i=1}^{n} \frac{1}{\sqrt{2 \pi \sigma^{2}}} \exp \left[-\frac{\varepsilon_{i}^{2}}{2 \sigma^{2}}\right] \\
& \propto \prod_{i=1}^{n} \frac{1}{\sigma} \exp \left[-\frac{\varepsilon_{i}^{2}}{2 \sigma^{2}}\right] \\
&=\frac{1}{\sigma^{n}} \cdot \exp \left[-\frac{\sum_{i=1}^{n} \varepsilon_{i}^{2}}{2 \sigma^{2}}\right]
\end{aligned}
$$

Likelihood of $\phi$ is equal to the likelihood of $\sigma^2$:

$$
\ell(\phi)=\ell\left(\sigma^{2}\right)=\phi^{\frac{n}{2}} \operatorname{exp}\left[-\phi \frac{\sum_{i=1}^{n} \varepsilon_{i}^{2}}{2}\right]
$$

## 1a
$$
\begin{aligned}
\pi_{1}\left(\sigma^{2}\right) & \sim \ell\left(\sigma^2\right) \cdot p(\sigma^2) \\
&=\ell\left(\delta^{2}\right) \cdot 1 \\
&=l\left(\sigma^{2}\right) \\
& = \frac{1}{\sigma^{n}} \cdot \exp \left[-\frac{\sum_{i=1}^{n} \varepsilon_{i}^{2}}{2 \sigma^{2}}\right] \\
& \sim \operatorname{IG}\left(\frac{n-1}{2}, \frac{1}{2} \sum_{i=1}^{n} \varepsilon_{i}^{2}\right)
\end{aligned}
$$

## 1b

$$
\begin{aligned}
\pi_{2}(\phi) & \propto \ell(\phi) \cdot p(\phi) \\
&=\ell(\phi) \cdot 1 \\
&=\ell(\phi) \\
& \sim \phi^{\frac{n}{2}} \exp \left[-\phi \frac{\sum_{i=1}^{n} \varepsilon_{i}^{2}}{2}\right] \\
& \sim G\left(\frac{n}{2}+1, \frac{\sum \varepsilon_{i}^{2}}{2}\right)
\end{aligned}
$$

# Problem 2
## 2.a
According to Equation 3.7 from the script, stationary distribution is defined as follows:
$$\pi=\pi P$$

Matrix $P$ is known to us.
Let's write $\pi$ as: $$\pi = [x  y  z]$$

Now let's find $x,y,z$ that satisfy equation 3.7.

- equation 3.7
- multiply left vector with matrix P and write both sides as colomn vectors
- factor $x,y$ and $z$ out of the brackets


$$[x \quad y \quad z] P=[x \quad y \quad z]$$
\begin{equation}
\begin{array}{l}

{\left[\begin{array}{lll}
0,5 x+0,3 y+0,2 z \\
0,4 x+0,4 y+0,3 z \\
0,1 x+0,3 y+0,5 z
\end{array}\right]} =\left[\begin{array}{l}
x \\
y \\
z
\end{array}\right] \\
{\left[\begin{array}{ccc}
-0,5 & 0,3 & 0,2 \\
0,4 & -0,6 & 0,3 \\
0,4 & 0,3 & -0,5
\end{array}\right]\left[\begin{array}{l}
x \\
y \\
z
\end{array}\right]=\left[\begin{array}{c}
0 \\
0 \\
0
\end{array}\right]}
\end{array}
\end{equation}

The probabilities have to sum up to 1, so we also have that $x+y+z=1$.
Together we get this system of linear equations.

\begin{equation}
\left[\begin{array}{ccc}
-0,5 & 0,3 & 0,2 \\
0,4 & -0,6 & 0,3 \\
0,4 & 0,3 & -0,5 \\
1 & 1 & 1
\end{array}\right]\left[\begin{array}{l}
x \\
y \\
z
\end{array}\right]=\left[\begin{array}{l}
0 \\
0 \\
0 \\
1
\end{array}\right]
\end{equation}

Solving this gets us the following result:
$$ x = 21/62, y = 32/62, z = 18/62 $$
$$ \pi = \frac{1}{62}[21 \quad 32 \quad 18] $$

## 2.b

Define transition kernels $p(1,\cdot)$,$p(2,\cdot)$,$p(3,\cdot)$

```{r}
p1 <- c(0.5, 0.4, 0.1)
p2 <- c(0.3, 0.4, 0.3)
p3 <- c(0.2, 0.3, 0.5)
```

```{r}
set.seed(42)
n <- 10^5
```

Sample the $\theta^{(0)}$ state uniformly from {1,2,3}.
```{r}
t <- numeric(n+1)
t[1] <- sample(c(1,2,3), size=1, prob = c(1,1,1))
```

```{r}
for (i in 2:(n+1)){
    if (t[i-1]==1){ p <- p1 }
    else if (t[i-1]==2){ p <- p2 }
    else { p <- p3 }
    t[i] <- sample(c(1,2,3), size=1, prob = p)
}
```

```{r}
rel_freq <- table(t) / n
rel_freq
```

Compare this to the analytical results:

```{r}
c(21/62, 23/62, 18/62)
```

```{r}
binded <- rbind(rel_freq, c(21/62, 23/62, 18/62))
barplot(binded, beside=T, main="
    Black - simulation
    Grey - analytical results")
```
