---
title: "Sampling in R"
author: "Claudia Czado"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document: 
    theme: paper
    toc: yes
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(continue = TRUE, warning = FALSE, message = FALSE,fig.width = 6, out.width = "70%",
                      fig.asp = 0.618, fig.retina = 3, fig.align = 'center')
knitr::opts_knit$set(global.par = TRUE)

```




## Required R packages

- rafalib
- mosaic
- dplyr
- ggplot2

```{r load, include=FALSE}
require(rafalib)
require(mosaic)
require(dplyr)
require(ggplot2)
```



## Flipping coins 

using the R packages

- mosaic, 
- dplyr and
- ggplot2

### Flipping a single fair coin
```{r}
set.seed(2017)
do(1) * rflip(1)
```

### Flipping 6 times 10 fair coins independently
```{r}
do(6) * rflip(10)
```

### Flipping 5000 times 5 fair coins independently
```{r}
simGuesses <- do(5000) * rflip(5)
```

Sorting the 5000 5 coin flips by the number of heads
```{r}
simGuesses %>% 
  group_by(heads) %>%
  summarize(count = n())
```


### Creating a histogram of the number of heads resulting from the 5000 flips

```{r, fig.width=4, out.width="50%"}
ggplot(data = simGuesses, mapping = aes(x = heads)) +
  geom_histogram(binwidth = 1, color = "white")

```

This histogram does not reflect the disceteness of the number of heads

### Histogram accounting for disceteness of the number of heads


```{r,fig.height=4, out.width="50%"}
ggplot(data = simGuesses, mapping = aes(x = factor(heads))) +
  geom_bar()
```

This histogram has the shape of a standard normal density, which is justified by the central limit theorem.
Adapted from Section 8.4 of [moderndiver](https://ismayc.github.io/moderndiver-book/8-sampling.html)


## Sampling from standard univariate distributions 

The function

- *runif()* samples from a uniform distribution

- *rbinom()* samples from a binomial distribution

- *rpois()* samples from a Poisson distribution

- *rgamma()* samples from a Gamma distribution

- *rexp()* samples from an exponential distribution

- *rnorm()* samples from a Normal distribution

### Sampling from the uniform and the normal distribution

We now sample from some distributions for several sample sizes


```{r}
data.unif100<-runif(100,min=0,max=1)
data.unif200<-runif(200,min=0,max=1)
data.unif500<-runif(500,min=0,max=1)
data.unif1000<-runif(1000,min=0,max=1)

data.norm100<-rnorm(100,mean=0,sd=1)
data.norm200<-rnorm(200,mean=0,sd=1)
data.norm500<-rnorm(500,mean=0,sd=1)
data.norm1000<-rnorm(1000,mean=0,sd=1)

data.exp100<-rexp(100,rate=1)
data.exp200<-rexp(200,rate=1)
data.exp500<-rexp(500,rate=1)
data.exp1000<-rexp(1000,rate=1)

```

### Histograms of the sampled uniformly distributed values

```{r, fig.height=7,fig.width=6}
bigpar(2,2)
hist(data.unif100,xlab="x",main="n=100")
hist(data.unif200,xlab="x",main="n=200")
hist(data.unif500,xlab="x",main="n=500")
hist(data.unif1000,xlab="x",main="n=1000")


```

### Histograms of the sampled exponentially distributed values

```{r, fig.height=7,fig.width=7}
bigpar(2,2)
hist(data.exp100,xlab="x",main="n=100")
hist(data.exp200,xlab="x",main="n=200")
hist(data.exp500,xlab="x",main="n=500")
hist(data.exp1000,xlab="x",main="n=1000")


```

### Histograms of the sampled normally distributed values

```{r, fig.height=7,fig.width=7}
bigpar(2,2)
hist(data.norm100,xlab="x",main="n=100")
hist(data.norm200,xlab="x",main="n=200")
hist(data.norm500,xlab="x",main="n=500")
hist(data.norm1000,xlab="x",main="n=1000")
```

### QQ plots of the normally distributed samples


```{r, fig.width=14}
bigpar(1,3)
qqnorm(data.norm200,main="n=200")
qqline(data.norm200,col=2)
qqnorm(data.norm500,main="n=500")
qqline(data.norm500,col=2)
qqnorm(data.norm1000,main="n=1000")
qqline(data.norm1000,col=2)

```

###  Further R functions for distributions

Note that

- \text{runif()} samples from the uniform distribution
- \text{punif()} gives the distribution function of the uniform distribution
- \text{dunif()} gives the density function of the uniform distribution
- \text{qunif()} gives the quantile function of the uniform distribution

Similarly for all the other distributions

###  Probability integral transforms (PIT) of normal and exponential samples

```{r}
udata.exp100<-pexp(data.exp100)
udata.exp200<-pexp(data.exp200)
udata.exp500<-pexp(data.exp500)
udata.exp1000<-pexp(data.exp1000)

udata.norm100<-pnorm(data.norm100)
udata.norm200<-pnorm(data.norm200)
udata.norm500<-pnorm(data.norm500)
udata.norm1000<-pnorm(data.norm1000)

```

### Histograms of PIT's of normally distributed samples

```{r,fig.width=7}
bigpar(2,2)
hist(udata.norm100,xlab="x",main="PIT(n=100)")
hist(udata.norm200,xlab="x",main="PIT(n=200)")
hist(udata.norm500,xlab="x",main="PIT(n=500)")
hist(udata.norm1000,xlab="x",main="PIT(n=1000)")
```


### Histograms of PIT's of exponentially distributed samples

```{r, fig.height=7,fig.width=7}
bigpar(2,2)
hist(udata.exp100,xlab="x",main="n=100")
hist(udata.exp200,xlab="x",main="n=200")
hist(udata.exp500,xlab="x",main="n=500")
hist(udata.exp1000,xlab="x",main="n=1000")
```
