---
title: "Gibbs sampling for the Poisson change point problem"
author: "Claudia Czado"
date: '`r format(Sys.time(), "%d %B, %Y")`'
header-includes:
   - \usepackage{framed}
output: 
  html_document: 
    theme: paper
    toc: yes
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(continue = TRUE, warning = FALSE, message = FALSE,fig.width = 6, out.width = "70%",
                      fig.asp = 0.618, fig.retina = 3, fig.align = 'center')
knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(cache = FALSE)
```



# Required R packages

- rafalib


```{r load, include=FALSE}
require(rafalib)
```




# Poisson change point problem

We assume we have 

- $Y_i \sim Pois(\lambda), i=1,\ldots,m$ and 
- $Y_i \sim Pois(\phi), i=m,\ldots,n$
where the change point $m$ is unknown. So we want to estimate
$\lambda, \phi$ and $m$.
- As prior distribution we assume $\lambda \sim Gamma(a,b)$ and $\phi \sim Gamma(\gamma,\delta)$

# Simulating data from the Poisson change point model

```{r}
sim.data.pois<-function(n = 50, a = 5, b = 1, ga = 1, de = 1)
{
	# generate data and plot it
	la <- rgamma(1, a, b)
	phi <- rgamma(1, ga, de)
	m <- sample(1:n, size = 1)
	y.m <- rpois(m, la)
	y.nm <- rpois(n - m, phi)
	y <- c(y.m, y.nm)
	par(mfrow = c(1, 1))
	plot(y)
	title(paste("lambda=", round(la, digits = 2), ", phi=", round(phi,
		digits = 2), ", m=", m))
	data <- list(n = n, y = c(y.m, y.nm), la.true = la, phi.true = phi,
		m.true = m, a = a, b = b, ga = ga, de = de)
	data
	}	
```


```{r,arsimdata}
data<-sim.data.pois(n = 50, a = 5, b = 1, ga = 1, de = 1)
```




# Gibbs sampling function

```{r}
gibbs.pois<-function(R = 100, data=data, m.start=25,
phi.start=1, la.start=1)
{
	la <- data$la.true
	phi <- data$phi
	m <- data$m
	n<-data$n
	a<-data$a
	b<-data$b
	ga<-data$ga
	de<-data$de
	y.m <- rpois(m, la)
	y.nm <- rpois(n - m, phi)
	y <- c(y.m, y.nm)
	# Gibbs Sampling
	la.vec <- rep(0, R)
	phi.vec <- rep(0, R)
	m.vec <- rep(0, R)
	prob <- rep(0, n)
	# starting  values
	m.old <- m.start
	la.old <- la.start
	phi.old <- phi.start
	for(i in 1:R) {
		la.new <- rgamma(1, a + sum(y[1:m.old]), b + m.old)
		if(m.old != n) {
			phi.new <- rgamma(1, ga + sum(y[(m.old + 1):n]), de +
				n - m.old)
		}
		else {
			phi.new <- rgamma(1, ga, de)
		}
		for(j in 1:(n - 1)) {
			temp <- (j * (phi.new - la.new)) + (log(la.new) * sum(
				y[1:j])) + (log(phi.new) * sum(y[(j + 1):n]))
			prob[j] <- temp
		}
		prob[n] <- (n * (phi.new - la.new)) + (log(la.new) * sum(y[
			1:n]))
		pmax <- max(prob)
		prob1 <- exp(prob - pmax)/sum(exp(prob - pmax))
		m.new <- sample(x = n, size = 1, replace = T, prob = prob1)
		#print(c(i, la.new, phi.new, m.new))
		la.vec[i] <- la.new
		phi.vec[i] <- phi.new
		m.vec[i] <- m.new
		m.old <- m.new
		la.old <- la.new
		phi.old <- phi.new
	}
	out <- list(data = data, phi = phi.vec, la = la.vec, m = m.vec,
	R = R,m.start=m.start,phi.start=phi.start,la.start=la.start,
	la.true=la,phi.true=phi,m.true=m)
	out
}

```

# Output function

```{r}
gibbs.plot<-function(out)
{
	# plots output from gibbs.poiss
	la <- out$la
	phi <- out$phi
	m <- out$m
	par(mfcol = c(3, 2))
	# time sequence plots
	plot(la, ylab = "lambda", xlab = "iteration", main = paste(
		"true lambda=", round(out$la.true, digits = 2)), type = 
		"l",sub=paste("start=",
		round(out$la.start,digits=2)))
	abline(h = out$la.true, lty = 2)
	plot(phi, ylab = "phi", xlab = "iteration", main = paste("true phi=",
		round(out$phi.true, digits = 2)), type = "l",
		sub=paste("start=",
		round(out$phi.start,digits=2)))
	abline(h = out$phi.true, lty = 2)
	plot(m, ylab = "m", xlab = "iteration", main = paste("true m=", round(
		out$m.true, digits = 2)), type = "l",sub=paste("start=",
		round(out$m.start,digits=2)))
	abline(h = out$m.true, lty = 2)
	title(paste("true m=", out$data$m.true))
	# density estimates
	hw <- 2 * (summary(la)[5] - summary(la)[2])
	plot(density(la, width = hw), xlab = "lambda", ylab = "probability",
		type = "l",main="")
	abline(v = out$data$la.true, lty = 2)
	title(paste("true lambda=", round(out$la.true, digits = 2)))
	hw <- 2 * (summary(phi)[5] - summary(phi)[2])
	plot(density(phi, width = hw), xlab = "phi", ylab = "probability",
		type = "l",main="")
	abline(v = out$phi.true, lty = 2)
	title(paste("true phi=", round(out$data$phi.true, digits = 2)))
	#hist(m, breaks = sort(unique(m)))
	barplot(table(m), names = as.character(sort(unique(m))))
	abline(v = out$m.true, lty = 2)
	title(paste("true m=", out$data$m.true))
}
```

# Running and examing the Gibbs sampler

```{r arsimout, fig.width=9,out.width='70%'}
out<-gibbs.pois(R=500,data)
gibbs.plot(out)

```



C. Czado, March 2020
