---
title: "Gibbs sampling with R2jags for the simple linear regression model with AR(1) distributed errors"
author: "Claudia Czado"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document: 
    theme: paper
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
```


# Required libraries
- R2jags: performs the Gibbs sampling
- rafalib: produces nicer output for multiple plots
- ggmcmc: tools for working with MCMC output
- xtable: produces latex tables from R output



```{r, include=FALSE}
require(R2jags)   # performs the Gibbs sampling
require(rafalib)  # produces nicer output for multiple plots
require(ggmcmc)   # tools for working with MCMC output
require(xtable)   # produces latex tables from R output
```


#  Generate data from a simple linear regression model with AR(1) disturbances 

```{r}
arlinsim<-function(n=200,b=c(1,.5),x=seq(-3,3,length=200),
                   sd.u=.5,rho=.8)
  {
  eps=rep(0,n)
  y=rep(0,n)
  eps[1]<-rnorm(1,sd.u)
  mu=rep(b[1],n)+(b[2]*x)
  for (t in 2:n){
       eps[t]=(rho*eps[t-1])+rnorm(1,mean=0,sd=sd.u)
	}
  y=mu+eps
  
  list(y=y,x=x,b.t=b,n=n,rho.t=rho,mu.t=mu,eps.t=eps,sd.u=sd.u)
}
set.seed(2020) # fix seed to get reproducibility
out<-arlinsim()
data_arlin=list(x=out$x,n=out$n,b0=c(0,0),
B0=matrix(c(0.00001, 0, 0,0.00001),nrow=2,ncol=2),y=out$y)
```


## Plot simulated data

```{r data_arsim,out.width="60%", fig.align = 'center'}
bigpar(3,1)
lab.tit<-paste("beta = (",out$b.t[1],",",out$b.t[2],")",", rho = ",out$rho.t,", sd.u = ",out$sd.u)
plot(1:out$n,out$eps.t,xlab="T",ylab="AR error",type="l",main=lab.tit)
plot(1:out$n,out$y,xlab="T",ylab="response",type="l")
plot(1:out$n,out$mu.t,xlab="T",ylab="mean",type="l")
```



# Set up bugs code


```{r}
arlin_model <-function(){
# initialization for t=1
   mean.y[1] <- b[1] + b[2]*x[1]
   y[1] ~ dnorm(mean.y[1], tau.u)
# t>1
   for(t in 2:n)
   {
      mean.y[t] <- b[1]*(1-rho) + b[2]*(x[t]-rho*x[t-1]) + rho*y[t-1]
      y[t] ~ dnorm(mean.y[t], tau.u)
   }

#  tau.u is the precision parameter
   sigma.u <- 1 / sqrt(tau.u)
#
# Prior specifications for
# precision
   tau.u ~ dgamma(0.05, 0.05)
# auto correlation
  rho ~ dunif(-1,1)
# regression parameter
  b[1:2] ~ dmnorm(b0[], B0[ , ])
}
```




# Run R2jags with single chain for 500 iterations with burnin of 100 iterations

```{r}
arlin_mcmc <- jags(data = data_arlin, inits=list(list(b=c(5,5),rho=0.0,tau.u=.01)),
             parameters.to.save = c("b","rho","sigma.u","tau.u"),
             n.iter = 500, 
             n.chains = 1,
             n.thin = 1,
             n.burnin = 0, # 100 iterations are run for adaption
             model.file = arlin_model)
```


## First check of output



```{r}
structure(arlin_mcmc) 
```

# Examining the output

## Trace plot in R2jags
```{r,out.width="70%", fig.align='center'}
bigpar(3,2)
traceplot(arlin_mcmc,mfrow=c(3,2))
```

## Examine MCMC output using ggmcmc

### Setup R2jags output to be used with ggmcmc
```{r}
# creates an mcmc object to be used in coda
arlin.mcmc <- as.mcmc(arlin_mcmc)
# imports MCMC samples to be used by ggs 
# graphical functions
arlin.ggs <- ggs(arlin.mcmc)
```

### Histograms of sampled values after burnin

```{r, out.width="70%", fig.align='center'}
ggs_histogram(arlin.ggs)
```

### Kernel density estimates of sampled values after burnin

```{r, out.width="70%", fig.align='center'}
ggs_density(arlin.ggs)
```

### Estimated autocorrelations among the chain

```{r acf_arsim, out.width="70%", fig.align='center'}
ggs_autocorrelation(arlin.ggs)
```

### Cross dependence between the sampled MCMC values

```{r pairs_arsim, out.width="70%", fig.align='center'}
ggs_pairs(arlin.ggs)
```

## Nicer layout

### Plot posterior density estimates and traces on a nicer layout
```{r, out.width="70%", fig.align='center'}
# traces and densities
mypar(3,4)
plot(arlin.mcmc,auto.layout=FALSE)
```

### Plot posterior traces on a nicer layout
```{r trace_arsim, out.width="70%", fig.align='center'}
# traces only
mypar(3,2)
plot(arlin.mcmc,density=FALSE,auto.layout = FALSE)
```

### Plot posterior density estimates on a nicer layout

```{r mar_den_arsim, out.width="70%", fig.align='center'}
# posterior density estimates only
mypar(3,2)
plot(arlin.mcmc,trace=FALSE,auto.layout = FALSE)
```

##  Examine DIC and other components

```{r}
names(arlin_mcmc$BUGSoutput) # see which output is available
arlin_mcmc$BUGSoutput$DIC
```


## MCMC summary statistics


```{r summary_arsim}
out<-summary(arlin.mcmc)
out
```


### Converting summary statistics to latex table

```{r summary_arsim_latex}
xtable(round(out$statistics,digits=3))
```


### Converting quantile results to latex table

```{r quantile_arsim_latex}
xtable(round(out$quantiles,digits=3))
```

C. Czado April 2020