---
title: "A Poisson Model for Claim Frequency Data"
author: "Claudia Czado"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document: 
    theme: paper
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
```



# Required libraries

- R2jags: performs the Gibbs sampling
- rafalib: produces nicer output for multiple plots
- ggmcmc: tools for working with MCMC output



```{r, include=FALSE}
require(R2jags)
library(rafalib)
require(ggmcmc)
```
# Data 

## Data notation

- x[i,j] is the number of claims for the ith group policy holder in jth policy year i=1..3 and j=1...5

- p[i,j] is the number on payroll for the ith group policy holder in jth policy year i=1..3 and j=1...5

## Input the data 

```{r}
data_claims=list( x = structure( 
              .Data = c( 9, 7, 6, 13,10, 
                                6, 4, 2, 8,  5, 
                             6, 8, 3, 4,  7 ), 
              .Dim = c( 3, 5 ) ), 

       p =  structure( 
              .Data = c( 280, 320, 265, 340,325, 
                                260, 275, 240, 265, 285, 
                                 267, 145, 120, 105, 115 ),  
              .Dim = c( 3, 5 ) ) ) 

data_claims
```


# Set up Bugs model to be used with \texttt{R2jags}

```{r}
claims_model <-function(){
for( j in 1 : 5 ) {
      for( i in 1 : 3 ) {
         x[i , j] ~ dpois(lambda[i , j])
      }
   }
for( j in 1 : 5 ) {
      for( i in 1 : 3 ) {
         lambda[i , j] <- p[i , j] * theta[i]
      }
   }
for( i in 1 : 3 ) {
      theta[i] ~ dgamma(alpha,beta)
   }
   alpha ~ dgamma(5,5)
   beta ~ dgamma(25,1)
}
```



# Run R2jags with single chain


```{r}
set.seed(2020)
claims_mcmc <- jags(data = data_claims, inits=list(list(alpha = 1,theta = c( 2, 2, 2 ), beta=1)),
             parameters.to.save =c("beta","alpha","theta","lambda"),
             n.iter = 1000, 
             n.chains = 1,
             n.thin = 1,
             n.burnin = 500, # 100 iterations are run for adaption
             model.file = claims_model)
```

## Examine R2jags output

```{r}
structure(claims_mcmc)
```


# Get MCMC output from R2jags to the right format


```{r}
claims.mcmc <- as.mcmc(claims_mcmc)
claims.ggs <- ggs(claims.mcmc)
# sampled values of first MCMC iteration
claims.mcmc[1,]
```

# Examine histograms of the sampled values

## Examine histograms of the sampled alpha values

```{r hist_alpha_claims, out.width="70%",fig.align='center'}
ggs_histogram(claims.ggs,family="alpha")

```

## Examine histograms of the sampled beta values

```{r hist_beta_claims, out.width="70%",fig.align='center'}
ggs_histogram(claims.ggs,family="beta")
```

## Examine histograms of the sampled theta values

```{r hist_theta_claims, out.width="70%",fig.align='center'}
ggs_histogram(claims.ggs,family="theta")
```


# Examine autocorrelations

## Examine autocorrolations of the sampled alpha values

```{r acf_alpha_claims, out.width="70%",fig.align='center'}
ggs_autocorrelation(claims.ggs,family="alpha")
```

## Examine autocorrolations of the sampled beta values

```{r acf_beta_claims, out.width="70%",fig.align='center'}
ggs_autocorrelation(claims.ggs,family="beta")
```

##  Examine autocorrolations of the sampled theta values

```{r acf_theta_claims, out.width="70%",fig.align='center'}
ggs_autocorrelation(claims.ggs,family="theta")
```

# Examine marginal posterior density estimates

## Examine marginal posterior density estimates for alpha


```{r mar_den_alpha_claims, out.width="70%",fig.align='center'}
ggs_density(claims.ggs,family="alpha")

```

##  Examine marginal posterior density estimates for beta

```{r mar_den_beta_claims, out.width="70%",fig.align='center'}
ggs_density(claims.ggs,family="beta")
```

## Examine marginal posterior density estimates for theta

```{r mar_den_theta_claims, out.width="70%",fig.align='center'}
ggs_density(claims.ggs,family="theta")
```




# Cross correlations for theta


```{r,out.width='60%', fig.align='center'}
ggs_pairs(claims.ggs,family="theta")
```

# Traces and density plots 
## Plot traces and posterior density  for alpha,beta and deviance
```{r,out.width="70%", fig.align='center'}
mypar(3,2)
plot(claims.mcmc[,1:3],auto.layout=FALSE)
```


## Plot traces and posterior density for theta
```{r,out.width="70%", fig.align='center'}
mypar(3,2)
plot(claims.mcmc[,19:21],auto.layout=FALSE)
```


## Plot posterior density estimates and traces for alpha, beta and theta
```{r trace_den_claims,fig.width=4, out.width="30%",fig.align='center'}
mypar(5,2)
plot(claims.mcmc[,c(1:2,19:21)],auto.layout=FALSE)
```


## Plot posterior density estimates and traces for DIC
```{r DIC_claims, out.width="60%",fig.align='center'}
# traces and densities
mypar(1,2)
plot(claims.mcmc[,3],auto.layout=FALSE)
```
##  Plot posterior density estimates for lambda
```{r den_lambda_claims, out.width="70%",fig.align='center'}
mypar(3,5)
plot(claims.mcmc[,4:18],trace=FALSE,auto.layout=FALSE,cex.main=1)
```


## Trace plots of sampled lambda


```{r trace_lambda_claims, out.width="70%",fig.align='center'}
mypar(3,5)
plot(claims.mcmc[,4:18],density=FALSE,auto.layout=FALSE,cex.main=1)
```


# MCMC summary statistics

```{r summary_arsim, out.width="70%",fig.align='center'}
out<-summary(claims.mcmc)
out
```
