---
title: "HMC analysis of a linear regression with autoregressive error using rtan applied to the approval rates of President Reagan"
author: "Claudia Czado"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document: 
    theme: paper
    toc: yes
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(cache = TRUE)
```


# Preliminaries


## Installing *Rtools*
- First one needs to install *Rtools*. Installation is described under 
*http://jtleek.com/modules/01_DataScientistToolbox/02_10_rtools/#1 *
- Downloads of *Rtools* you can find under *https://cran.r-project.org/bin/windows/Rtools/ *

- After *Rtools* is installed install in Rstudio the package *devtools*

## Installing R package *bayesjackman*

- In *Rstudio* type into the console window

*devtools::install_github("jrnold/jackman-bayes", subdir = "bayesjackman")*
  
## Required R packages

- bayesjackman (contains rstan codes for the examples of Simon Jackman)
- rafalib 
- rstan
- tidyverse


```{r load, include=FALSE}
require(rafalib)
require(bayesjackman)
require(rstan)
require(tidyverse)
```

# Package *bayesjackman*

- Under *https://jrnold.github.io/bugs-examples-in-stan/ * you can find 
a description of **Simon Jackman's Bayesian Model Examples in Stan**
by Jeffrey B. Arnold. 
- His work is licensed under the Creative Commons Attribution 4.0 International License
*https://creativecommons.org/licenses/by /4.0/ *
- The following analysis is based on his code.

# Approval rates of President Reagan

The data set *ReaganApproval* is contained in the R package *bayesjackman* and gives the *approval rating* of President Reagan over 96 months together with the corresponding *unemployment* and *inflation rates*.

```{r}
data("ReaganApproval", package = "bayesjackman")
head(ReaganApproval)
```

# Exploring the data

```{r}
summary(ReaganApproval)
```

```{r reagan_data,fig.height=8, fig.width=6}
bigpar(3,1)
plot(1:96,ReaganApproval$app,type="l",ylab="app",xlab="month")
plot(1:96,ReaganApproval$infl,type="l",ylab="infl",xlab="month")
plot(1:96,ReaganApproval$unemp,type="l",ylab="unemp",xlab="month")
```

## Fitting a linear model

```{r}
r<-lm(app~infl+unemp, data=ReaganApproval)
summary(r)
```


## Checking residuals over time

The residuals are correlated over time
```{r}
round(cor(resid(r)[1:95],resid(r)[2:96]),digits=2)
```



```{r resid_ar, fig.height=5,fig.width=6}
plot(resid(r)[1:95],resid(r)[2:96])
abline(0,1,lty=2)
```

# Model formulation for the linear model with autoregressive disturbances

Note the marginal model for the first observation, and the conditioning on the lagged observation for months 2 through 96. A uniform prior over the stationary (-1,1) interval is employed for the residual AR(1) parameter.

- $y_i=\mu_i+\epsilon_i+\theta \epsilon_{i-1}$

- $\mu_i=\alpha+x^\top\beta$ and $\epsilon_{i} \sim Normal(0,\sigma^2)$, i.i.d.
for $i \in 1,.,N$. 

Weakly informative priors for each parameter are used,


- $\alpha \sim Normal(0,10),\beta_k \sim Normal(0,2.5),k \in 1,.,K$

- $\sigma \sim HalfCauchy(0,5),\theta=2\theta^* - 1, \theta^* \sim Beta(1,1)$.


# Setup the data for running *rstan*

```{r}
reagan_data <- within(list(), {
  y <- ReaganApproval$app
  N <- length(y)
  X <- model.matrix(~ 0 + infl + unemp, data = ReaganApproval) %>% scale()
  K <- ncol(X)
  alpha_loc <- 0
  alpha_scale <- 10
  beta_loc <- rep(0, K)
  beta_scale <- rep(2.5 * sd(y), K)
  sigma_scale <- 5 * sd(y)
  theta_a <- 1
  theta_b <- 1
})
```

# Setup for a linear regression model with autoregressive errors to be used in *rstan*

In the file *regar1.stan* the stan model formulation for a linear regression model with autoregressive error is given. This takes some time when you run it the first time.

```{r, cache=TRUE}
mod_regar1 <- stan_model("regar1.stan")
```

To see the stan code type:
```{r}
mod_regar1
```

# Run a MCMC using Hamiltonian Monte Carlo using *rstan*

The function *sampling* performs the sampling. Some of its defaults
are

- chains=4

- iter=2000

- warmup=floor(iter/2)

- thin=1

- init='random'

- show_messages=TRUE

For details you can find with the help function.

```{r,eval=FALSE}
help("sampling,stanmodel-method")
```


```{r}
reagan_fit <- sampling(mod_regar1, data = reagan_data)
```

# Summary output of the HMC samples

```{r}
out_reagan<-round(summary(reagan_fit, par = c("alpha", "beta", "theta", "sigma"))$summary,digits=3)
out_reagan
```

# Trace plots of the HMC sample

```{r trace_reagan,fig.width=8,fig.height=6}
traceplot(reagan_fit,par = c("alpha", "beta", "theta", "sigma"),inc_warmup=TRUE)


```

# Marginal posterior density estimates 

```{r mar_den_reagan,fig_width=8,fig_height=6}
stan_dens(reagan_fit,par = c("alpha", "beta", "theta", "sigma"),separate_chains=TRUE)

```


# Estimated posterior autocorrelations

```{r acf_reagan,fig_width=8,fig_height=6}
stan_ac(reagan_fit,par = c("alpha", "beta", "theta", "sigma"),separate_chains=TRUE)
```



